Modal type theory as based on the intuitionistic modal logic IEL$^{-}$

### 2018-05-08 article.tex
update in article

\documentclass[a4paper]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{bussproofs}
\usepackage{mathtools}
\usepackage{verbatim}
\usepackage{dsfont}
\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage[all, 2cell]{xy}
\usepackage[all]{xy}
\newtheorem{theorem}{Theorem}
\newtheorem{prop}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{defin}{Definition}
\newtheorem{ex}{Example}
\newtheorem{col}{Corollary}
\newtheorem{con}{Consequence}
\usepackage{listings} 		% for source code
\date{}
\title{Modal type theory based on the intuitionistic epistemic logic}

@@

\begin{abstract}
Modal intuitionistic epistemic logic IEL$^{-}$ was proposed by S.Artemov and T. Protopopescu as the formal
foundation for the intuitionistic theory of knowledge. We construct a modal simply typed lambda-calculus
which is Curry-Howard isomorphic to IEL$^{-}$ as formal theory of calculations with applicative functors in
functional programming languages like Haskell or Idris.
\end{abstract}

@@

  Modal intutionistic epistemic logic IEL was proposed by S. Artemov and T. Proropopescu \cite{Artemov}. IEL
  provides the epistimology and the theory of knowledge as based on BHK-semantics of intuitionistic logic.
  IEL$^{-}$  is a variant of IEL, that corresponds to intuitionistic belief.
  Informally, $\Box A$ denotes that $A$ is verified intuitionistically.

@@

  Functional programming languages such as Haskell \cite{Haskell}, Idris \cite{Idris} or Purescript \cite{Purs}
  have special type classes\footnote{Type class in Haskell is a general interface for
  special group of datatypes.} for calculations with container types like \verb"Functor" and
  \verb"Applicative" \footnote{Reader may read more about container types in the Haskell standard library documentation\cite{Base} or in the next one textbook
  \cite{Miran}}:

@@

  By \emph{container} (or \emph{computational context}) type we mean some type-operator $f$, where $f$ is a
  ``function'' from $*$ to $*$: type operator takes a simple type (ща kind $*$) and returns another
  simple type of kind \verb"*". For more detailed description of the type system with kinds used in
  Haskell see \cite{Morten}.

@@

  We investigate the relationship between intuitionistic epistemic logic IEL$^{-}$ and applicative programming with side-effects
  by constructing the type system (which is called $\lambda_{\text{IEL}^{-}}$) which is Curry-Howard isomorphic to $IEL^{-}$.
  So we will consider IEL$^{-}$ modality as an arbitrary applicative functor and we prove that obtained type system
  is sound and complete for applicative functor on cartesian closed category (using the categorical definition proposed by Paterson \cite{Cons}).

@@

  \begin{minipage}{0.5\textwidth}
    \begin{flushleft}
  	\begin{prooftree}
    	\AxiomC{$\Gamma \vdash A$}
    	\RightLabel{$\Box_I$}
    	\UnaryInfC{$\Gamma \vdash \Box A$}
  \end{prooftree}
    \end{flushleft}
  \end{minipage}
  \begin{minipage}{0.5\textwidth}
    \begin{flushright}
  	\begin{prooftree}
    \AxiomC{$\Gamma \vdash \Box A_1, \dots, \Gamma \vdash \Box A_n $}
    \AxiomC{$A_1,\dots,A_n \vdash B$}
    \BinaryInfC{$\Gamma \vdash \Box B$}
  	\end{prooftree}
    \end{flushright}
  \end{minipage}
  \end{defin}

The first rule allows to derive co-reflextion. The second modal rule is a counterpart of $\Box_I$
rule in natural deduction calculus for constructive K (see \cite{ModalLa}).

@@

$\begin{array}{lll}
(1) &\bigwedge \Gamma \rightarrow \Box A_1, \dots, \bigwedge \Gamma \rightarrow \Box A_1 & \text{assumption} \\
(2) &\bigwedge \Gamma \rightarrow \bigwedge \limits_{i = 1}^{n} \Box A_i & \text{IEL$^{-}$ theorem} \\
(3) &\bigwedge \limits_{i = 1}^{n} \Box A_i \rightarrow \Box \bigwedge \limits_{i = 1}^{n} A_i & \text{IEL$^{-}$ theorem} \\
(4) &\bigwedge \Gamma \rightarrow \Box \bigwedge \limits_{i = 1}^{n} A_i & \text{from (1), (2) and transitivity} \\
(5) &\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B& \text{assumption} \\
(6) &(\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B) \rightarrow \Box (\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B)& \text{co-reflection}\\
(7) &\Box (\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B)& \text{from (4), (5) and MP} \\
(8) &\Box \bigwedge \limits_{i = 1}^{n} A_i \rightarrow \Box B & \text{from (6) and normality} \\
(9) &\bigwedge \Gamma \rightarrow \Box B & \text{from (3), (7) and transitivity}
\end{array}$

@@

\begin{proof}
  Straightforward derivation of modal axioms in $\text{NIEL}^{-}$. We consider this derivation below using terms.
\end{proof}

@@

  At the next step we build the typed lambda-calculus based on the NIEL$^{-}$ with implication and dijunction
by proof-assingment in rules. Obtained fragment is equivalent to IEL$^{-}$ without axioms for negation and disjunction.

@@

\begin{enumerate}
  \item $FV(x) = \{ x \}$;
  \item $FV(\lambda x. M) = FV(M) \setminus \{ x\}$;
  \item $FV(M N) = FV(M) \cup FV(N);$
  \item $FV(\langle M,N \rangle) = FV(M) \cup FV(N)$;
  \item $FV(\pi_i M) \subseteq FV(M)$, $i \in \{ 1, 2\}$;
  \item $FV(\text{\bf box } M) = FV(M)$;
  \item $FV({\bf let \: box} \: \vec{x} = \vec{M} {\: \bf in \:} N) = \bigcup \limits_{i = 1}^n FV(M), \text{where $n = |\vec{M}|$}$.
\end{enumerate}
  \end{defin}

@@

1) The derivation ends in

\begin{prooftree}
  \AxiomC{$\Gamma \vdash \vec{M} : \Box \vec{A}$}
  \AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
  \RightLabel{$\text{let}_{\Box}$}
  \BinaryInfC{$\Gamma \vdash {\bf let \: box \:} \vec{x} = \vec{M} {\: \bf in \: } N : \Box B$}
\end{prooftree}

By IH $\Delta \vdash \vec{M} : \Box \vec{A}$, so $\Delta \vdash {\bf let \: box \:} \vec{x} = \vec{M} {\: \bf in \: } N : \Box B$.

2)-3) Similary.

@@

1) If $\Gamma \vdash {\bf let \: box \:} \vec{x}, y, \vec{z} = \vec{M}, {\bf let \: box \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P}  {\: \bf in \:} R : \Box B$,
  then $\Gamma \vdash {\bf let \: box \:} \vec{x}, \vec{w}, \vec{z} = \vec{M}, \vec{N}, \vec{P} \: { \bf in } \: R [y := Q] : \Box B$ by rule 4).

2) Let $\Gamma \vdash {\bf let \: box \:} x = M {\: \bf in \: } x : \Box A$, then $\Gamma \vdash M : \Box A$ by rule 9).

See \cite{ModalK}.

3) The derivation ends in

\begin{prooftree}
\AxiomC{$\Gamma \vdash {\bf box \:} \vec{M} : \Box \vec{A}$}
\AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
\BinaryInfC{$\Gamma \vdash {\bf let \: box \:} \vec{x} = {\bf box \:} \vec{M} {\: \bf in \:} N : \Box B$}
\end{prooftree}

So $\Gamma \vdash \vec{M} : \vec{A}$ by Lemma 4 and $\Gamma \vdash N [\vec{x} := \vec{M}] : B$ by Lemma 4, part 3.

Then we can transform this into the following derivation:

\begin{prooftree}
\AxiomC{$\Gamma \vdash N [\vec{x} := \vec{M}] : B$}
\RightLabel{$\Box_I$}
\UnaryInfC{$\Gamma \vdash {\bf box \:} N [\vec{x} := \vec{M}] : \Box B$}
\end{prooftree}

4) The derivation ends in

\begin{prooftree}
\AxiomC{$\vdash M : A$}
\UnaryInfC{$\vdash {\bf let \: box \:} \underline{\quad} = \underline {\quad} {\: \bf in \:} M : \Box A$}
\end{prooftree}

So, if $\vdash M : A$, then $\vdash {\bf box \:} M : \Box A$.

@@

  Let us define the transformation from $\lambda_{\text{IEL}^{-}}$ into the simple type theory with
  $\to$, $\times$ and natural number type $\mathbb{N}$ \footnote{Strong normalization for stronger system was shown here \cite{Girard}} with additional typing and reduction rules:

@@

  \begin{defin} Interpretation of terms:
    \begin{enumerate}
      \item $x \in \mathbb{V} \Rightarrow |x| = x$;
      \item $|\lambda x. M| = \lambda x. |M|$;
      \item $|(M N)| = |M| |N|$;
      \item $|\langle M, N \rangle| = \langle |M|, |N| \rangle$;
      \item $|\pi_i M| = \pi_i |M|$, $i \in \{ 1, 2\}$;
      \item $|{\bf box \:} M| = \langle 0, |M| \rangle$;
      \item $|{\bf let \: box \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} M| = \langle 0, |M| \rangle$;
      \item $|{\bf let \: box \:} x = N {\: \bf in \:} M| = \langle \pi_1 |N| , |M| [x := \pi_2 |N|] \rangle$
      \item $|{\bf let \: box \:} \vec{x} = \vec{N} {\: \bf in \:} M| = \langle \sum \limits_{i = 1}^n \pi_1 |N|, |M| [\vec{x} := \pi_2 \vec{N}] \rangle$
    \end{enumerate}
  \end{defin}

@@

Also we may consider four critical pairs which are used in local confluence proof for lambda-calculus based on {\bf IK} and \cite{ModalK}.

@@

\begin{proof}
  By induction on the structure of $M$.
  Case with ${\bf let \: box \:} \vec{x} = \vec{M} {\: \bf in \:} N$ was considered by Kakutani \cite{ModalK} \cite{ModalK1}.

@@

  \xymatrix{
    && (\mathcal{F}A \otimes_{\mathcal{D}} \mathcal{F}B) \otimes_{\mathcal{D}} \mathcal{F}C \ar[d]_{(A \varoast B) \otimes_{\mathcal{D}} id_{\mathcal{F}B}}
    \ar[rr]^{\alpha^{\mathcal{D}}_{{\mathcal{F}A, \mathcal{F}B, \mathcal{F}C}}} && \mathcal{F}A \otimes_{\mathcal{D}} (\mathcal{F}B \otimes_{\mathcal{D}} \mathcal{F}C) \ar[d]^{id_{\mathcal{F}A} \otimes_{\mathcal{D}} \ast_{B,C}}\\
    && \mathcal{F}(A \otimes_{\mathcal{C}} B) \otimes_{\mathcal{D}} \mathcal{C} \ar[d]_{(A \otimes_{\mathcal{C}} B) \varoast C} && \mathcal{F}A \otimes_{\mathcal{D}} \mathcal{F}(B \otimes_{\mathcal{C}} C) \ar[d]^{A \varoast (B \otimes_{\mathcal{C}} C) }\\
    && \mathcal{F}((A \otimes_{\mathcal{C}} B) \otimes_{\mathcal{C}} C) \ar[rr]_{\mathcal{F}(\alpha^{\mathcal{C}}_{A,B,C})}&& \mathcal{F}(A \otimes_{\mathcal{C}} (B \otimes_{\mathcal{C}} C))
  }

@@

By default we will consider some monoidal functor on cartesian closed category below.

@@

\begin{defin} Semantical translation from $\lambda_{\text{IEL}^{-}}$ to some cartesian closed category $\mathcal{C}$ with an applicative functor $\boxdot$:

@@

\begin{small}
  \begin{prooftree}
    \AxiomC{$[\![\Gamma \vdash \vec{M} : \Box \vec{A}]\!] = \langle [\![M_1]\!],\dots, [\![M_n]\!] \rangle : [\![\Gamma]\!] \rightarrow \prod \limits_{i=1}^n \boxdot [\![A_i]\!]$}
    \AxiomC{$[\![\vec{x} : \vec{A} \vdash N : B]\!] = [\![N]\!] : \prod \limits_{i=1}^n [\![A_i]\!] \rightarrow [\![B]\!]$}
    \BinaryInfC{$[\![\Gamma \vdash {\bf let \: box \:} \vec{x} = \vec{M} {\: \bf in \: } M : \Box B]\!] = \boxdot ([\![N]\!]) \circ \ast_{[\![A_1]\!],\dots,[\![A_n]\!]} \circ \langle [\![M_1]\!],\dots, [\![M_n]\!] \rangle : [\![\Gamma]\!] \rightarrow \boxdot [\![B]\!]$}
  \end{prooftree}
\end{small}
\end{enumerate}
\end{defin}

@@

$\begin{array}{lll}
&[\![\Gamma \vdash ({\bf let \: box} \: \vec{x} = \vec{M} {\: \bf in \:} N) [\vec{y} := \vec{P}] : \Box B]\!] =& \\
&\quad\quad\quad\quad\quad\quad\quad \text{Substitution} &\\
&[\![\Gamma \vdash {\bf let \: box} \: \vec{x} = (\vec{M} [\vec{y} := \vec{P}]) {\: \bf in \:} N : \Box B]\!] =& \\
&\quad\quad\quad\quad\quad\quad\quad  \text{Interpretation for $let_{\Box}$} \\
&\boxdot ([\![N]\!]) \circ ([\![A_1]\!] \ast \dots \ast [\![A_n]\!]) \circ [\![\Gamma \vdash (\vec{M} [\vec{y} := \vec{P}]) : \Box \vec{A}]\!] =& \\
&\quad\quad\quad\quad\quad\quad\quad \text{Induction hypothesis}& \\
&\boxdot ([\![N]\!]) \circ ([\![A_1]\!] \ast \dots \ast [\![A_n]\!]) \circ ([\![\vec{M}]\!] \circ \langle [\![P_1]\!],\dots,[\![P_n]\!]\rangle) = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{Associativity of composition}& \\
&(\boxdot ([\![N]\!]) \circ ([\![A_1]\!] \ast \dots \ast [\![A_n]\!]) \circ [\![\vec{M}]\!]) \circ \langle [\![P_1]\!],\dots,[\![P_n]\!]\rangle = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{Interpretaion}& \\
&[\![\Gamma \vdash {\bf let \: box} \: \vec{x} = \vec{M} {\: \bf in \:} N : \Box B]\!] \circ \langle [\![P_1]\!],\dots,[\![P_n]\!]\rangle&
\end{array}$

@@

$\begin{array}{lll}
&[\![\vdash {\bf let \: box \:} \underline{\quad} = \underline {\quad} {\: \bf in \:} M : \Box A]\!] = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{By interpretation}& \\
&\boxdot ([\![M]\!]) \circ u_{\mathds{1}} = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{By the definition of an applicative functor}& \\
&\boxdot {K}([\![M]\!]) \circ \eta_{\mathds{1}} = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{By naturality of $\eta$} & \\
&\eta_{[\![A]\!]} \circ [\![M]\!] = &\\
&\quad\quad\quad\quad\quad\quad\quad \text{By interpretation} & \\
&[\![\vdash {\bf box \:} M : \Box A]\!]&
\end{array}$
\end{proof}

@@

\begin{enumerate}
  \item $\text{fmap }(g \circ f) = \text{fmap }(g) \circ \text{fmap }(f)$;
  \item $\text{fmap }(id_{\hat{A}}) = id_{\boxdot}\hat{A}$.
\end{enumerate}
\end{lemma}

@@

\begin{enumerate}
  \item $\eta: Id_{\mathcal{C}(\lambda)} \Rightarrow \boxdot$, s. t. $\forall \hat{A} \in Ob_{\mathcal{C}(\lambda)}$, $\eta_{\hat{A}} = [x, {\bf box \:} x] \in Hom_{\mathcal{C}(\lambda)}(\hat{A}, \boxdot \hat{A})$;
  \item $\hat{A} \varoast \hat{B} : \boxdot \hat{A} \times \boxdot \hat{B} \to \boxdot (\hat{A} \times \hat{B})$, s. t. $\forall \hat{A}, \hat{B} \in Ob_{\mathcal{C}(\lambda)}, \ast_{\hat{A},\hat{B}} = [p, {\bf let \: box \:} x,y = \pi_1 p, \pi_2 p {\: \bf in \:} \langle x, y \rangle] \in Hom_{\mathcal{C}(\lambda)}(\boxdot \hat{A} \times \boxdot \hat{B}, \boxdot (\hat{A} \times \hat{B}))$.
\end{enumerate}
\end{defin}

Implementation for $\ast$ in our term model is the instance of ${\text{let}_{\Box}}$-rule:

>>>>

\documentclass[a4paper]{article}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{bussproofs}
\usepackage{mathtools}
\usepackage{verbatim}
\usepackage{dsfont}
\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage[all, 2cell]{xy}
\usepackage[all]{xy}
\usepackage{authblk}
\newtheorem{theorem}{Theorem}
\newtheorem{prop}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{defin}{Definition}
\newtheorem{ex}{Example}
\newtheorem{col}{Corollary}
\newtheorem{con}{Consequence}
\usepackage{listings} 		% for source code
\date{}
\author[1]{Daniel Rogozin}
\affil[1]{Moscow State University, Moscow, Russia}
\affil[1]{Serokell O\"{U}, Tallinn, Estonia}
\title{Modal type theory as based on the intuitionistic modal logic IEL$^{-}$}

@@

\begin{abstract}
Modal intuitionistic epistemic logic IEL$^{-}$ was proposed by S. Artemov and T. Protopopescu as the formal
foundation for the intuitionistic theory of knowledge. We construct a modal simply typed lambda-calculus
which is Curry-Howard isomorphic to IEL$^{-}$ as the formal theory of calculations with applicative functors in
functional programming.
\end{abstract}

@@

  Modal intutionistic epistemic logic IEL was proposed by S. Artemov and T. Proropopescu \cite{Artemov}. IEL
  provides the epistimology and the theory of knowledge as based on BHK-semantics of intuitionistic logic.
  IEL$^{-}$  is a variant of IEL, that corresponds to intuitionistic belief.
  Informally, $\Box A$ denotes that the agent believes in $A$.

@@

  Functional programming languages such as Haskell, Idris or Purescript
  have special type classes \footnote{Type class in Haskell is a general interface for
  special group of datatypes.} for calculations with type operators like \verb"Functor" and
  \verb"Applicative" \footnote{Reader may read more about these types in the Haskell standard library documentation.}:

@@

  By \emph{container} (or \emph{computational context}) type we mean some type operator $f$, where $f$ is a
  ``function'' from $*$ to $*$: type operator takes a simple type (ща kind $*$) and returns another
  simple type of kind \verb"*". For more detailed description of the type system with kinds used in
  Haskell see \cite{Morten}.

@@

  We investigate the relationship between intuitionistic epistemic logic IEL$^{-}$ and applicative programming with side-effects
  by constructing the type system (which is called $\lambda_{\text{IEL}^{-}}$) which is Curry-Howard isomorphic to $IEL^{-}$.
  So we will consider IEL$^{-}$ modality as an arbitrary applicative functor and we prove that the obtained type system
  is sound and complete for an applicative functor on cartesian closed category
  (using the categorical definition proposed by Paterson \cite{Cons}).

@@

  \begin{center}
    \begin{prooftree}
      \AxiomC{$ $}
      \RightLabel{\bf{ax}}
      \UnaryInfC{$\Gamma, A \vdash A$}
    \end{prooftree}
  \end{center}

  \begin{minipage}{0.5\textwidth}
    \begin{flushleft}
    \begin{prooftree}
      \AxiomC{$\Gamma, A \vdash B$}
      \RightLabel{$\to_I$}
      \UnaryInfC{$\Gamma \vdash A \to B$}
    \end{prooftree}
    \begin{prooftree}
      \AxiomC{$\Gamma \vdash A$}
      \AxiomC{$\Gamma \vdash B$}
      \RightLabel{$\land_I$}
      \BinaryInfC{$\Gamma \vdash A \land B$}
    \end{prooftree}
  	\begin{prooftree}
    	\AxiomC{$\Gamma \vdash A$}
    	\RightLabel{${\Box_I}_1$}
    	\UnaryInfC{$\Gamma \vdash \Box A$}
  \end{prooftree}
    \end{flushleft}
  \end{minipage}
  \begin{minipage}{0.5\textwidth}
    \begin{flushright}
    \begin{prooftree}
      \AxiomC{$\Gamma \vdash A \to B$}
      \AxiomC{$\Gamma \vdash A$}
      \RightLabel{$\to_E$}
      \BinaryInfC{$\Gamma \vdash B$}
    \end{prooftree}
    \begin{prooftree}
      \AxiomC{$\Gamma \vdash A_1 \land A_2$}
      \RightLabel{$\land_E, i \in {1,2}$}
      \UnaryInfC{$\Gamma \vdash A_i$}
    \end{prooftree}
  	\begin{prooftree}
    \AxiomC{$\Gamma \vdash \Box A_1, \dots, \Gamma \vdash \Box A_n $}
    \AxiomC{$A_1,\dots,A_n \vdash B$}
    \RightLabel{${\Box_I}_2$}
    \BinaryInfC{$\Gamma \vdash \Box B$}
  	\end{prooftree}
    \end{flushright}
  \end{minipage}
  \end{defin}

Without loss of generality we will consider the calculus without inference rule for disjunction and negation.
Obtained calculus is equivalent to Hilbert-style IEL$^{-}$ without axioms for negation and disjunction.

The first modal rule allows to derive co-reflextion. The second modal rule is a counterpart of $\Box_I$
rule in natural deduction calculus for constructive K (see \cite{ModalLa}).

@@

$\begin{array}{lll}
(1) &\bigwedge \Gamma \rightarrow \Box A_1, \dots, \bigwedge \Gamma \rightarrow \Box A_1 & \text{assumption} \\
(2) &\bigwedge \Gamma \rightarrow \bigwedge \limits_{i = 1}^{n} \Box A_i & \text{IEL$^{-}$ theorem} \\
(3) &\bigwedge \limits_{i = 1}^{n} \Box A_i \rightarrow \Box \bigwedge \limits_{i = 1}^{n} A_i & \text{IEL$^{-}$ theorem} \\
(4) &\bigwedge \Gamma \rightarrow \Box \bigwedge \limits_{i = 1}^{n} A_i & \text{from (2), (3) and transitivity} \\
(5) &\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B& \text{assumption} \\
(6) &(\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B) \rightarrow \Box (\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B)& \text{co-reflection}\\
(7) &\Box (\bigwedge \limits_{i = 1}^{n} A_i \rightarrow B)& \text{from (5), (6) and MP} \\
(8) &\Box \bigwedge \limits_{i = 1}^{n} A_i \rightarrow \Box B & \text{from (7) and normality} \\
(9) &\bigwedge \Gamma \rightarrow \Box B & \text{from (4), (8) and transitivity}
\end{array}$

@@

\begin{proof}
  Straightforward derivation of modal axioms in $\text{NIEL}^{-}$. We shall consider these derivations using terms below.
\end{proof}

@@

  At the next step we build the typed lambda-calculus based on the NIEL$^{-}$.
by proof-assingment in rules.

@@

\begin{enumerate}
  \item $FV(x) = \{ x \}$;
  \item $FV(\lambda x. M) = FV(M) \setminus \{ x\}$;
  \item $FV(M N) = FV(M) \cup FV(N);$
  \item $FV(\langle M,N \rangle) = FV(M) \cup FV(N)$;
  \item $FV(\pi_i M) = FV(M)$, $i \in \{ 1, 2\}$;
  \item $FV(\text{\bf box } M) = FV(M)$;
  \item $FV({\bf let \: box} \: \vec{x} = \vec{M} {\: \bf in \:} N) = \bigcup \limits_{i = 1}^n FV(M), \text{where $n = |\vec{M}|$}$.
\end{enumerate}
  \end{defin}

@@

Induction on the derivation of $\Gamma \vdash M : A$.

@@

\begin{enumerate}

\item If $\Gamma \vdash {\bf let \: box \:} \vec{x}, y, \vec{z} = \vec{M}, {\bf let \: box \: } \vec{w} = \vec{N} {\bf \: in \: } Q, \vec{P}  {\: \bf in \:} R : \Box B$,
  then $\Gamma \vdash {\bf let \: box \:} \vec{x}, \vec{w}, \vec{z} = \vec{M}, \vec{N}, \vec{P} \: { \bf in } \: R [y := Q] : \Box B$ by rule 4).
\item Let $\Gamma \vdash {\bf let \: box \:} x = M {\: \bf in \: } x : \Box A$,
then $\Gamma \vdash M : \Box A$ by rule 9). \\ See \cite{ModalK}.
\item The derivation ends in
\begin{prooftree}
\AxiomC{$\Gamma \vdash {\bf box \:} \vec{M} : \Box \vec{A}$}
\AxiomC{$\vec{x} : \vec{A} \vdash N : B$}
\BinaryInfC{$\Gamma \vdash {\bf let \: box \:} \vec{x} = {\bf box \:} \vec{M} {\: \bf in \:} N : \Box B$}
\end{prooftree}
So $\Gamma \vdash \vec{M} : \vec{A}$ by Lemma 4 and $\Gamma \vdash N [\vec{x} := \vec{M}] : B$ by Lemma 4, part 3.
Then we can transform this into the following derivation:
\begin{prooftree}
\AxiomC{$\Gamma \vdash N [\vec{x} := \vec{M}] : B$}
\RightLabel{$\Box_I$}
\UnaryInfC{$\Gamma \vdash {\bf box \:} N [\vec{x} := \vec{M}] : \Box B$}
\end{prooftree}
\item The derivation ends in
\begin{prooftree}
\AxiomC{$\vdash M : A$}
\UnaryInfC{$\vdash {\bf let \: box \:} \underline{\quad} = \underline {\quad} {\: \bf in \:} M : \Box A$}
\end{prooftree}
So, if $\vdash M : A$, then $\vdash {\bf box \:} M : \Box A$.
\end{enumerate}
\end{proof}

@@

  Let us define the transformation from $\lambda_{\text{IEL}^{-}}$ into the simple type theory with
  $\to$, $\times$ and natural number type $\mathbb{N}$ with additional typing and reduction rules
  \footnote{Strong normalization for stronger system was shown here \cite{Girard}} :

@@

  \begin{defin} Interpretation of terms:
    \begin{enumerate}
      \item $x \in \mathbb{V} \Rightarrow |x| = x$;
      \item $|\lambda x. M| = \lambda x. |M|$;
      \item $|(M N)| = |M| |N|$;
      \item $|\langle M, N \rangle| = \langle |M|, |N| \rangle$;
      \item $|\pi_i M| = \pi_i |M|$, $i \in \{ 1, 2\}$;
      \item $|{\bf box \:} M| = \langle 0, |M| \rangle$;
      \item $|{\bf let \: box \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} M| = \langle 0, |M| \rangle$;
      \item $|{\bf let \: box \:} x = N {\: \bf in \:} M| = \langle \pi_1 |N| , |M| [x := \pi_2 |N|] \rangle$
      \item $|{\bf let \: box \:} \vec{x} = \vec{N} {\: \bf in \:} M| = \langle \sum \limits_{i = 1}^n \pi_1 |N|, |M| [\vec{x} := \pi_2 |\vec{N}|] \rangle$
    \end{enumerate}
  \end{defin}

@@

Also we may consider four critical pairs which are considered in
confluence proof for lambda-calculus as based on {\bf IK} \cite{ModalK}.

@@

  By induction on the structure of $M$.
  Case with ${\bf let \: box \:} \vec{x} = \vec{M} {\: \bf in \:} N$ was considered by Kakutani \cite{ModalK} \cite{ModalK1}.

@@

  \xymatrix{
    && (\mathcal{F}A \otimes_{\mathcal{D}} \mathcal{F}B) \otimes_{\mathcal{D}} \mathcal{F}C \ar[d]_{(A \varoast B) \otimes_{\mathcal{D}} id_{\mathcal{F}B}}
    \ar[rr]^{\alpha^{\mathcal{D}}_{{\mathcal{F}A, \mathcal{F}B, \mathcal{F}C}}} && \mathcal{F}A \otimes_{\mathcal{D}} (\mathcal{F}B \otimes_{\mathcal{D}} \mathcal{F}C) \ar[d]^{id_{\mathcal{F}A} \otimes_{\mathcal{D}} B \varoast C}\\
    && \mathcal{F}(A \otimes_{\mathcal{C}} B) \otimes_{\mathcal{D}} \mathcal{C} \ar[d]_{(A \otimes_{\mathcal{C}} B) \varoast C} && \mathcal{F}A \otimes_{\mathcal{D}} \mathcal{F}(B \otimes_{\mathcal{C}} C) \ar[d]^{A \varoast (B \otimes_{\mathcal{C}} C) }\\
    && \mathcal{F}((A \otimes_{\mathcal{C}} B) \otimes_{\mathcal{C}} C) \ar[rr]_{\mathcal{F}(\alpha^{\mathcal{C}}_{A,B,C})}&& \mathcal{F}(A \otimes_{\mathcal{C}} (B \otimes_{\mathcal{C}} C))
  }

@@

By default we will consider an arbitrary monoidal functor on cartesian closed category below.

@@

\begin{defin} Semantical translation from $\lambda_{\text{IEL}^{-}}$ to some cartesian closed category $\mathcal{C}$ with an applicative functor $\langle \mathcal{C}, \boxdot, \eta \rangle$:

@@

\begin{small}
  \begin{prooftree}
    \AxiomC{$[\![\Gamma \vdash \vec{M} : \Box \vec{A}]\!] = \langle [\![M_1]\!],\dots, [\![M_n]\!] \rangle : [\![\Gamma]\!] \rightarrow \prod \limits_{i=1}^n \boxdot [\![A_i]\!]$}
    \AxiomC{$[\![\vec{x} : \vec{A} \vdash N : B]\!] = [\![N]\!] : \prod \limits_{i=1}^n [\![A_i]\!] \rightarrow [\![B]\!]$}
    \BinaryInfC{$[\![\Gamma \vdash {\bf let \: box \:} \vec{x} = \vec{M} {\: \bf in \: } M : \Box B]\!] = \boxdot ([\![N]\!]) \circ ([\![A_1]\!] \varoast \dots \varoast [\![A_n]\!]) \circ \langle [\![M_1]\!],\dots, [\![M_n]\!] \rangle : [\![\Gamma]\!] \rightarrow \boxdot [\![B]\!]$}
  \end{prooftree}
\end{small}
\end{enumerate}
\end{defin}

Interpretation for ${\bf let \: box\:}$-rule is similar to interpretation for $\Box$-rule in term calculus for intutionistic {\bf K} \cite{Paiva}.

@@

$\begin{array}{lll}
&[\![\Gamma \vdash ({\bf let \: box} \: \vec{x} = \vec{M} {\: \bf in \:} N) [\vec{y} := \vec{P}] : \Box B]\!] =& \\
&\quad\quad\quad\quad\quad\quad\quad \text{Substitution} &\\
&[\![\Gamma \vdash {\bf let \: box} \: \vec{x} = (\vec{M} [\vec{y} := \vec{P}]) {\: \bf in \:} N : \Box B]\!] =& \\
&\quad\quad\quad\quad\quad\quad\quad  \text{Interpretation for $let_{\Box}$} \\
&\boxdot ([\![N]\!]) \circ ([\![A_1]\!] \varoast \dots \varoast [\![A_n]\!]) \circ [\![\Gamma \vdash (\vec{M} [\vec{y} := \vec{P}]) : \Box \vec{A}]\!] =& \\
&\quad\quad\quad\quad\quad\quad\quad \text{Induction hypothesis}& \\
&\boxdot ([\![N]\!]) \circ ([\![A_1]\!] \varoast \dots \varoast [\![A_n]\!]) \circ ([\![\vec{M}]\!] \circ \langle [\![P_1]\!],\dots,[\![P_n]\!]\rangle) = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{Associativity of composition}& \\
&(\boxdot ([\![N]\!]) \circ ([\![A_1]\!] \varoast \dots \varoast [\![A_n]\!]) \circ [\![\vec{M}]\!]) \circ \langle [\![P_1]\!],\dots,[\![P_n]\!]\rangle = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{Interpretaion}& \\
&[\![\Gamma \vdash {\bf let \: box} \: \vec{x} = \vec{M} {\: \bf in \:} N : \Box B]\!] \circ \langle [\![P_1]\!],\dots,[\![P_n]\!]\rangle&
\end{array}$

@@

$\begin{array}{lll}
&[\![\vdash {\bf let \: box \:} \underline{\quad} = \underline {\quad} {\: \bf in \:} M : \Box A]\!] = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{By interpretation}& \\
&\boxdot ([\![M]\!]) \circ u_{\mathds{1}} = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{By the definition of an applicative functor}& \\
&\boxdot ([\![M]\!]) \circ \eta_{\mathds{1}} = & \\
&\quad\quad\quad\quad\quad\quad\quad \text{By naturality of $\eta$} & \\
&\eta_{[\![A]\!]} \circ [\![M]\!] = &\\
&\quad\quad\quad\quad\quad\quad\quad \text{By interpretation} & \\
&[\![\vdash {\bf box \:} M : \Box A]\!]&
\end{array}$
\end{proof}

@@

\begin{enumerate}
  \item $\boxdot(g \circ f) = \boxdot g \circ \boxdot f$;
  \item $\boxdot (id_{\hat{A}}) = id_{\boxdot}\hat{A}$.
\end{enumerate}
\end{lemma}

@@

\begin{enumerate}
  \item $\eta: Id_{\mathcal{C}(\lambda)} \Rightarrow \boxdot$, s. t. $\forall \hat{A} \in Ob_{\mathcal{C}(\lambda)}$, $\eta_{\hat{A}} = [x, {\bf box \:} x] \in Hom_{\mathcal{C}(\lambda)}(\hat{A}, \boxdot \hat{A})$;
  \item $\hat{A} \varoast \hat{B} : \boxdot \hat{A} \times \boxdot \hat{B} \to \boxdot (\hat{A} \times \hat{B})$, s. t. $\forall \hat{A}, \hat{B} \in Ob_{\mathcal{C}(\lambda)}, \hat{A} \varoast \hat{B} = [p, {\bf let \: box \:} x,y = \pi_1 p, \pi_2 p {\: \bf in \:} \langle x, y \rangle] \in Hom_{\mathcal{C}(\lambda)}(\boxdot \hat{A} \times \boxdot \hat{B}, \boxdot (\hat{A} \times \hat{B}))$.
\end{enumerate}
\end{defin}

Implementation for $\varoast$ in our term model is the instance of ${\text{let}_{\Box}}$-rule:
